---
title: ""
author: "Davide Lorino"
date: "2023-06-08"
output: 
  github_document:
    toc: true
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(knitr)
library(tibble)
```

# Most Relevant Explanation

First introduced in <i>Yuan, C. et al. (2011), <a href src="https://arxiv.org/abs/1401.3893">"Most Relevant Explanations in Bayesian Networks"</a>, Journal of Artiﬁcial Intelligence Research, pp. 309-352</i> - the Most Relevant Expalantion in a Bayesian network is the partial instantiation of nodes that maximizes the Generalized Bayes Factor. 

# Generalized Bayes Factor

The Generalized Bayes Factor is introduced by Yuan et al in the aforementioned paper and it is also a deterministic function on the Weight of Evidence (WoE) first introduced by Alan Turing. The WoE is the logarithm of the GBF (see Good, I. J. (1985), <a src= "https://www.cs.tufts.edu/~nr/cs257/archive/jack-good/weight-of-evidence.pdf">"Weight of Evidence: A Brief Survey"</a>, Bayesian Statistics 2, pp. 249-270, for more general discussion of history and concepts).

## Exhaustive Search Complexity

The challenge of calculating the Generalized Bayes Factor is that it requires the posterior odds to be calculated for every possible partial instantiation of nodes in the network. Partial instantiations of Bayesian Networks is an NP-hard problem where the number of partial instantiations grows super-exponentially with each node and node-state in the network.

In concrete terms, if we assume a highly optimized algorithm that can calculate Generalized Bayes Factors for 10 partial instantiations in the network per second, these would be the running times if we calculated Generalized Bayes Factors on Bayesian networks using an exhaustive search algorithm:

```{r echo=FALSE}
df <- tibble(
  `Number of Nodes` = c(5, 5, 10, 20),
  `States per Node` = c(5, 10, 10, 10),
  `Total Combinations` = c("5^5 = 3125", "10^5 = 100,000", "10^10 = 10,000,000,000", "10^20 = 100,000,000,000,000,000,000"),
  `Expected Runtime` = c("~0.005 minutes", "~0.167 minutes", "~31.7 years", "~3.17 billion years")
)

knitr::kable(df)
```


As you can see, the expected runtime increases dramatically as the number of nodes and states per node increase. 
This is why exhaustive search becomes computationally infeasible for even fairly small networks of only 10 nodes.


## Hierarchical Beam Search

The Hierarchical Beam Search algorithm is proposed by Changhe Yuan and Xiaoyuan Zhu in <i><a href src="https://www.sciencedirect.com/science/article/pii/S1570868316300854">'Hierarchical beam search for solving most relevant explanation in Bayesian networks, Journal of Applied Logic, Volume 22, 2017, Pages 3-13'</i></a> as an approximate inference algorithm that performs remarkably well on a suite of Bayesian network benchmarks and works by applying multi-stage scan-and-prune logic, pruning candidate hypotheses that don't improve the Generalized Bayes Factor and gorwing a blacklist of hypotheses from which to cull successive candidate hypotheses until every Generalized Bayes Factor in the search space has either been explicitly evaluated or pruned. This reduction in the search space + pruning criteria combination is what makes it possible for this algorithm to enable reliable inference but also be computationally feasible on even large networks. 

# The MRE R Package

At the time of writing on June 8th 2023 there is no other open source implementation for calculating the Most Relevant Explanation in a Bayesian network, either exhaustively or through the proposed Hierarchical Beam Search. The `MRE` R package implements the Hierarchical Beam Search algorithm proposed by Yuan and Zhu for calculating the Most Relevant Explanation in a Bayesian network.

# Example Usage

```{r message=FALSE}
# load mre
library(mre)

# create a synthetic dataset of age, gender and income
set.seed(42)
n <- 10000
age <- factor(sample(c("young", "adolescent", "old"), n, replace = TRUE, prob = c(0.33, 0.33, 0.34)))
gender <- factor(sample(c("male", "nonbinary", "female"), n, replace = TRUE, prob = c(0.33, 0.33, 0.34)))

# define probabilities for income based on age and gender
income_probs <- function(age_val, gender_val) {
  if (age_val == "old" & gender_val == "male") {
    return(c("low" = 0.01, "high" = 0.99))
  } else if (age_val == "young" & gender_val == "female") {
    return(c("low" = 0.99, "high" = 0.01))
  } else {
    return(c("low" = 0.99, "high" = 0.01))
  }
}

# generate income variable
income <- factor(mapply(function(a, g) {
  sample(c("low", "high"), 1, replace = TRUE, prob = income_probs(a, g))
}, age, gender))

# join age, gender and income into a dataframe
raw_data <- data.frame(age, gender, income)

# define the targets that you want explanations for 
target_var = "income"
target_val = "high"

# run
mre_solution <- mre::mre_hierarchical_beam_search(
  x = raw_data, 
  target_variable = target_var, 
  target_value = target_val
  )

# show the candidate hypotheses with the greatest Generalized Bayes Factors
knitr::kable(mre_solution)
```

















